Towards Compiler-Agnostic Performance in Finite-Difference Codes

### 2015-08-07 docs/papers/shallow_performance/shallow_perf.tex
Add fuller description of transformations and way in which they were chosen

The \psykal approach separates code into three layers; the Algorithm
layer, the PSy layer and the Kernel layer. While this approach is
general, we are currently applying it to Atmosphere and Ocean models
written in Fortran where domain decomposition is typically performed
in the latitude-longitude direction, leaving columns of elements on
each domain-decomposed partition.

@@

%{\bf Auto-generated BC code by reasoning about meta-data. But then,
%  that brings us on to meta-data...}

@@

{\bf RF: Explain that we gradually apply transformations which restructure the psykal code to end up with the same optimised structure as the original code. Explain the reason behind the ordering and what each does - I think you said it was the order that you thought would make the biggest difference. If you can base it on performance analysis then all the better but I don't think that is required.}

{\bf Performance analysis as motivation for the code
  tranformations. More detail on each transformation.}

Beginning with the vanilla \psykal version, we then apply a series of
code transformations while obeying the \psykal separation of concerns,
{\it i.e.} optimisation is restricted to the middle, {PS}y layer and leaves
the kernel and algorithm layers unchanged. The aim of these
optimisations is to recover, as much as is possible, the structure and
thus performance of the original version of the code. The
transformations we have performed are as follows:
\begin{enumerate}

\item Pass explicit array bounds to middle layer (rather than using
  assumed-size array arguments)

\item Move kernel routines into same module as middle layer

\item Loop fusion

\item In-lined field-copy operation

\item Fused field-copy operation

\item Manually in-line kernel bodies into middle-layer code

\item Fully fuse first loop (only the outer loop of this loop nest was
  fused in the first pass)

\item Make the array/loop bounds explicitly constant for the
      duration of the time-stepping loop.
\end{enumerate}

>>>>

The \psykal approach attempts to address the problems described in the
previous section. It separates code into three layers; the Algorithm
layer, the PSy layer and the Kernel layer. The approach has been
developed in the GungHo project~\cite{GungHo}, which is creating a new
Dynamical core for the Met Office, and its design has been influenced
by earlier work on OP2~\cite{OP2,PYOP2}.

While this approach is general, we are currently applying it to
Atmosphere and Ocean models written in Fortran where domain
decomposition is typically performed in the latitude-longitude
direction, leaving columns of elements on each domain-decomposed
partition.

@@

Excluding the application of boundary conditions, the code has three
main sections: computation of mass fluxes, vorticity and sea-surface
height; computation of the pressure and velocity components; and
finally, smoothing of the fields in time and copying of the field
values at the next time-step to be those at the current
time-step. Profiling of the original code showed that these three
sections each account for approximately one third of the run-time.

@@

Beginning with the vanilla \psykal version, we then apply a series of
(cumulative) code transformations while obeying the \psykal separation
of concerns, {\it i.e.} optimisation is restricted to the middle,
{PS}y layer and leaves the kernel and algorithm layers unchanged. The
aim of these optimisations is to recover, as much as is possible, the
structure and thus performance of the original version of the code. We
began with two transformations that affect all of the kernels. After
this, we determined the transformations to perform by comparing the
timing profiles of the \psykal and original versions of the code. We
then tackled the sections resposible for the largest slowdowns and
iterated this process. While doing this we also made heavy use of the
optimisation reports produced by the compilers and how these changed
when moving from the original to the \psykal version of the code.

The transformations we have performed are as follows:
\begin{enumerate}

\item Specify array bounds in the middle layer using module variables
  (rather than using assumed-size array arguments);

\item Move all kernel subroutines into the same module as as that
  containing the middle/PSy layer (module inlining);

\item Loop fusion: all neighbouring loop nests were fully fused (where
  possible) with the exception of the first loop nest where it was
  found that fusing only the outer loop was more performant.

\item Manually in-line (kernel inline) the three field-copy operations
  performed as part of the time-update section;

\item Loop fuse the three field-copy operations with the
  time-smoothing loop nest;

\item Manually in-line all kernel bodies into the middle-layer
  subroutine (kernel inlining);

\item Fully fuse the first loop nest (only the outer loop of this loop
  nest was fused in the first pass);

\item Make the array/loop bounds explicitly constant for the duration
  of the time-stepping loop (pass as arguments to the middle layer
  rather than using module variables).

\end{enumerate}

In addition to these transformations we also applied two
transformations that were suggested by the IBM compiler team. The
first of these manually blocked the outer loop of the time-smoothing
loop nest using a factor of four. The second effectively undoes our 3rd
transformation by fissioning the first loop nest back into three
separate loop nests.
